Conceptual and methodological problems with bias detection and avoidance in natural language processing

### 2021-06-23 thesis/_book/_main.tex
Adds logs

\documentclass[12pt,]{book}

@@

            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage[left=3.5cm, right=2.5cm, top=2.5cm, bottom=2.5cm, asymmetric,
includeheadfoot]{geometry}

@@

{
\hypersetup{linkcolor=black}
\setcounter{tocdepth}{5}
\tableofcontents
}
\chapter{Introduction}\label{introduction}

@@

One can find various definitions trying to capture what bias and
fairness actually are. With the choice of the definition, implications
for the real-life applications may change as well. Mehrabi, Morstatter,
Saxena, Lerman, \& Galstyan
(\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) mark out that there
exist different types of biases such as historical bias, representation
bias or measurement bias (the list is long). This indicates how complex
the issue of bias is. Without the proper understanding and awareness of
the problem, people are prone to unconsciously sustain the bias. Mehrabi
et al. (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) also
distinguish different types of discrimination, some of them will be
briefly described. By protected attributes we mean those qualities,
traits or characteristics that one should not discriminate against.
Direct discrimination occurs when protected attributes of individuals
explicitly result in non-favorable outcomes toward them. Such
discrimination takes place for example when one uses the information
about someone's religious beliefs to evaluate whether they are competent
for the work they apply for. In contrast, in indirect discrimination
individuals appear to be treated equally, but anyway they end up being
treated unjustly due to the hidden effects of biases towards their
protected attributes. An example of such situation is when in a job
advert there is a requirement of 10 years experience instead of the list
of a specific type of experience and knowledge. This job advert can
discriminate indirectly against young people who can have the required
skills, but not yet so much work experience. Systemic discrimination
takes place when policies, customs or behaviors that result from certain
culture or organizational structure lead to discrimination against some
groups of people. Finally, statistical discrimination consists in
relying on group statistics to judge a person belonging to a given
group. The topic of discrimination is entangled with another concept;
fairness. It is essential to grasp some concepts of fairness to take
them into consideration while designing implementation of some machine
learning model. In Mehrabi et al.
(\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) one may notice that
depending on the context and application different definitions may be
applied.

There is a considerable amount of literature available on the topic of
bias detection and mitigation in NLP models. Bolukbasi, Chang, Zou,
Saligrama, \& Kalai (\protect\hyperlink{ref-Bolukbasi2016Man}{2016})
focus on gender biases that may be observable in the representation of
job occupations and gender in terms of numerical values. The authors
apply cosine similarity measurement to investigate the phenomenon where
(the vectors corresponding to) words related to jobs that are
stereotypically associated with a given gender are in fact in the model
situated closer to this gender. They also use analogy tasks to evaluate
if a bias is present in a word embedding model. They check analogies by
comparing pairs of word vectors. For example they search for the word
complementing the puzzle: man is to doctor as woman is to \ldots{}?
First they subtract the word ``man'' from the word ``woman''\footnote{This
  is pointwise subtraction of vectors, but for the simplicity we will
  often talk about words while really meaning vectors. What is meant
  should be clear from the context.} and then they search for the ranked
list of other words pairs that have similar vector difference. They also
include in the formula a threshold to ensure that the resulting pairs
could not be randomly picked.

However, as Nissim, Noord, \& Goot
(\protect\hyperlink{ref-Nissim2019Fair}{2019}) points out, there are
some limitations to this approach. In practice, most of analogies
implementations do not return any input words. This means that it does
not make sense to expect the algorithm to return the same profession for
both woman and man. Therefore, this method of bias detection seems
limited. Other problems are related for example to the choice of pairs
and words that are used to detect the presence of discrimination, as it
is often subjective and without proper justification. Additionally, the
choice of the parameter used in Bolukbasi et al.
(\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) formula to ensure that
word pairs are not picked by random, is also not justified and changing
it drastically influences the results.

Islam, Bryson, \& Narayanan
(\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) touch upon the
topic of biases regarding race and gender. They apply knowledge from
well-known psychological studies such as Implicit Association Test to
investigate the relation between human stereotypical thinking and
model-learnt biases to discover a close relationship between these two.
For the evaluation they use Word Embedding Association Test (WEAT) and
the Word Embedding Factual Association Test (WEFAT).

Manzini, Lim, Tsvetkov, \& Black
(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) propose a
novel way of using cosine similarity to verify if the assumed
resemblance between protected words and harmful attributes exists. They
develop an approach that enables them to measure the bias for a class
(such as gender, religion or race) and express the final result with a
single metric. Their approach will be analyzed in details in the next
chapter.

Islam et al. (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) refer
to Implicit Association Test (Greenwald et al., 1998) that measures the
strength of associations between concepts or stereotypes by measuring
human reaction time for special tasks. Humans naturally exhibit some
biases which do not always cause social concern. One can imagine the
intuitive associations between for example insects and flowers, and the
feelings of pleasantness or unpleasantness. In general, people would
rather associate flowers than insects with feeling pleasant, and this
preference in a sense is a bias or prejudice in some direction. However,
this type of preference does not cause an uproar and is a rather morally
neutral case. This example is rather used to show that the methodology
used in IAT makes sense. Islam et al.
(\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) were able to
obtain similar results to the ones that measured reaction time with the
use of WEAT metric. Unfortunately, further studies discover other biases
and prejudices that directly influence the quality of other people's
lives and therefore they should be taken care of.

@@

\section{Cosine similarity and
distance}\label{cosine-similarity-and-distance}

@@

\begin{align} \tag{Sim}
\mathsf{cosineSimilarity}(A,B) & = \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert}
\end{align}

Cosine similarity is considered a proper tool for this operation as its
result has a clear connection to geometry and at least for a low number
of dimensions may be easily interpreted. Using this scale, one can
compare vector similarities in a fairly clear manner. When the vectors
are aligned perpendicularly to each other, their similarity equals 0
(which is the same as the cosine of 90 degrees). This tells us that the
similarity between the vectors is small. As the angle between vectors
decreases, cosine similarity approaches one, which stands for the
greatest similarity. It is also possible to obtain negative cosine
similarity. If the value approaches -1, this intuitively means that the
words are contrary to each other.

@@

\section{Cosine distance in a one-class bias
detection}\label{cosine-distance-in-a-one-class-bias-detection}

Bolukbasi et al. (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) define
similarity between words as the outcome inner product of their
normalized vectors. They focus on examining what the geometry of a word
embedding is in regard to ``he'' and ``she'' words. In other words,
whether the similarity between those concepts and other words reflects
expected gender stereotypes. They test this hypothesis by investigating
whether there is a connection between word embeddings representing
certain professions and words referring to gender. They also evaluate
whether automatically produced analogies between words reflect the
stereotypes as well.

Here the gender bias of a word \(w\) is understood as its projection on
the gender direction
\(\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})\) (the
gender direction is the top principal compontent of ten gender pair
difference vectors). The underlying idea is that no bias is present if
non-explicitly gender words are in equal distance to both elements in
all explicitly gender pairs. Given the (ideally) gender neutral words
\(N\) and the gender direction \(g\) the direct direct gender bias is
defined as the average distance of the words in \(N\) from \(g\) (\(c\)
is a parameter determining how strict we want to be):

\begin{align}
\mathsf{directBias_c(N,g)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},g)\vert^c}{\vert N \vert }
\end{align}

@@

\section{Cosine distance in a multi-class bias
detection}\label{cosine-distance-in-a-multi-class-bias-detection}

Manzini et al.
(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) present a
different approach towards finding similarities between classes of
words. Cosine distance is used in the article as a measure to first
argue for the existence of multi-class bias and then to show how through
bias mitigation techniques the bias may be decreased.

They modify WEAT to a multi-class setting, introducing Mean Average
Cosine similarity as a measure of bias (in fact, in the paper they
report distances rather than similarities). Let
\(T = \{t_1, \dots, t_k\}\) be a class of protected word embeddings, and
let each \(A_j\in A\) be a set of attributes stereotypically associated
with a protected word). Then:

\begin{align}
S(t_i, A_j) & = \frac{1}{\vert A_j\vert}\sum_{a\in A_j}\mathsf{cos}(t,a) \\
MAC(T,A) & = \frac{1}{\vert T \vert \,\vert A\vert}\sum_{t_i \in T }\sum_{A_j \in A} S(t_i,A_j)
\end{align}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{religion <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
\KeywordTok{colnames}\NormalTok{(religion) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
                        \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
\NormalTok{religion}\OperatorTok{$}\NormalTok{wordClass <-}\StringTok{ }\KeywordTok{as.factor}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass)}
\KeywordTok{levels}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
\KeywordTok{head}\NormalTok{(religion)  }\OperatorTok{%>%}\StringTok{  }\KeywordTok{kable}\NormalTok{(}\DataTypeTok{format =} \StringTok{"latex"}\NormalTok{,}\DataTypeTok{booktabs=}\NormalTok{T,}
                      \DataTypeTok{linesep =} \StringTok{""}\NormalTok{,  }\DataTypeTok{escape =} \OtherTok{FALSE}\NormalTok{, }
                      \DataTypeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\OperatorTok{%>%}
\StringTok{                      }\KeywordTok{kable_styling}\NormalTok{(}\DataTypeTok{latex_options=}\KeywordTok{c}\NormalTok{(}\StringTok{"scale_down"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

Assuming for a moment that the value of multi-class cosine distance is
correct, one may question the interpretation. Manzini et al.
(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) summarize the
averages of cosine distance per group (gender, race, religion). For now
let us focus now on analyzing the values relating to religious biases.
Here is the relevant fragment of table:

\begin{longtable}[]{@{}ll@{}}
\toprule
Religion Debiasing & MAC\tabularnewline
\midrule
\endhead
Biased & 0.859\tabularnewline
Hard Debiased & 0.934\tabularnewline
Soft Debiased (\(\lambda\) = 0.2) & 0.894\tabularnewline
\bottomrule
\end{longtable}

@@

Using cosine similarity in high dimensions in word embeddings may also
be prone to the curse of dimensionality. According to Venkat
(\protect\hyperlink{ref-Venkat2018Curse}{2018}) there are reasons to
consider this phenomenon when searching for word similarities in higher
dimensions. An experiment is conducted that aims at showing how the
similarity values and variation change as the number of dimensions
increases. The hypothesis made in the paper states that two things will
happen as the number of dimensions increase. First, the effort required
to measure cosine similarity will be greater, and two, the similarity
between data will blur out and have less variation. The authors generate
random points with increasing number of dimensions where each dimension
of a data point is given a value between 0 and 1. Then they pick one
vector at random from each dimension class and calculate the cosine
similarity between the chosen vector and the rest of the data. Then they
check how the variation of values changes as the number of dimensions
increases. It seems like the more dimensions there are, the smaller the
variance and therefore it is less obvious how to interpret the resulting
cosine similarities. Maybe the scale should be adjusted to the number of
dimensions and variance so that it still gives us sensible information
about data. Yet, according to some articles cosine similarity in high
dimensions is not reliable enough as it may be the case that choosing
words at random may result in similar values as when picking them
consciously.

@@

We will use the choice of protected words and stereotypical predicates
used in Manzini et al.
(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}). This is a
decent point of departure, as not only we want to compare our method to
that of Manzini et al.
(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}), but also
because this data format is fairly general (as contrasted, say, with a
set up for binary stereotypes). Note also that the method we develop
here can fairly easily be run for different stereotypization patterns.
Let's start with explaining the method and its deployment using a
dataset obtained for the religion-related protected words.

Let's load, clean a bit and inspect the head of the religion dataset we
prepared. In order to obtain this dataset, we calculated the cosine
distance between each protected word and each word from both the
bias-related attribute groups, which were used in the original study,
and to neutral and human control attributes which we added as control
groups. For instance, for religion, the bias-related predicates (coming
from the original study in Manzini et al.
(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019})) include
muslim bias attributes, jew bias attributes, christian bias attributes
(see a list in the Appendix).

@@

\caption{\label{tab:religionTableHead}Head of the religion dataset.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lllrrl}
\toprule
protectedWord & wordToCompare & wordClass & cosineDistance & cosineSimilarity & connection\\
\midrule
judaism & violent & muslim & 0.7141939 & 0.2858061 & different\\
judaism & terrorist & muslim & 0.7461333 & 0.2538667 & different\\
judaism & dirty & muslim & 1.2002599 & -0.2002599 & different\\
judaism & uneducated & muslim & 0.7885469 & 0.2114531 & different\\
judaism & greedy & jewish & 1.0026172 & -0.0026172 & associated\\
judaism & cheap & jewish & 1.2323229 & -0.2323229 & associated\\
\bottomrule
\end{tabular}}
\end{table}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{religion}\OperatorTok{$}\NormalTok{con <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(religion}\OperatorTok{$}\NormalTok{connection)}
\NormalTok{religion}\OperatorTok{$}\NormalTok{pw <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(religion}\OperatorTok{$}\NormalTok{protectedWord)}
\NormalTok{religion}\OperatorTok{$}\NormalTok{pwFactor <-}\StringTok{ }\KeywordTok{factor}\NormalTok{(}\KeywordTok{paste0}\NormalTok{(religion}\OperatorTok{$}\NormalTok{protectedWord, religion}\OperatorTok{$}\NormalTok{connection))}
\NormalTok{religion}\OperatorTok{$}\NormalTok{pwIndex <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(religion}\OperatorTok{$}\NormalTok{pwFactor)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{source}\NormalTok{(}\StringTok{"../functions/cleanDataset.R"}\NormalTok{)}
\NormalTok{religion <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
\NormalTok{religion <-}\StringTok{ }\KeywordTok{cleanDataset}\NormalTok{(religion,}\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}
%
\documentclass[
  12pt,
]{book}
\usepackage{amsmath,amssymb}

@@

  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[left=3.5cm, right=2.5cm, top=2.5cm, bottom=2.5cm, asymmetric, includeheadfoot]{geometry}

@@

{
\hypersetup{linkcolor=}
\setcounter{tocdepth}{5}
\tableofcontents
}
\setstretch{1.5}
\hypertarget{introduction}{%
\chapter{Introduction}\label{introduction}}

@@

One can find various definitions trying to capture what bias and fairness actually are. With the choice of the definition, implications for the real-life applications may change as well. \protect\hyperlink{ref-Mehrabi2019Survey}{Mehrabi, Morstatter, Saxena, Lerman, \& Galstyan} (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) mark out that there exist different types of biases such as historical bias, representation bias or measurement bias (the list is long). This indicates how complex the issue of bias is. Without the proper understanding and awareness of the problem, people are prone to unconsciously sustain the bias. \protect\hyperlink{ref-Mehrabi2019Survey}{Mehrabi, Morstatter, Saxena, Lerman, \& Galstyan} (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) also distinguish different types of discrimination, some of them will be briefly described. By protected attributes we mean those qualities, traits or characteristics that one should not discriminate against. Direct discrimination occurs when protected attributes of individuals explicitly result in non-favorable outcomes toward them. Such discrimination takes place for example when one uses the information about someone's religious beliefs to evaluate whether they are competent for the work they apply for. In contrast, in indirect discrimination individuals appear to be treated equally, but anyway they end up being treated unjustly due to the hidden effects of biases towards their protected attributes. An example of such situation is when in a job advert there is a requirement of 10 years experience instead of the list of a specific type of experience and knowledge. This job advert can discriminate indirectly against young people who can have the required skills, but not yet so much work experience. Systemic discrimination takes place when policies, customs or behaviors that result from certain culture or organizational structure lead to discrimination against some groups of people. Finally, statistical discrimination consists in relying on group statistics to judge a person belonging to a given group. The topic of discrimination is entangled with another concept; fairness. It is essential to grasp some concepts of fairness to take them into consideration while designing implementation of some machine learning model. In \protect\hyperlink{ref-Mehrabi2019Survey}{Mehrabi, Morstatter, Saxena, Lerman, \& Galstyan} (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) one may notice that depending on the context and application different definitions may be applied.

There is a considerable amount of literature available on the topic of bias detection and mitigation in NLP models. \protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) focus on gender biases that may be observable in the representation of job occupations and gender in terms of numerical
values. The authors apply cosine similarity measurement to investigate the phenomenon where (the vectors corresponding to) words related to jobs that are stereotypically associated with a given gender are in fact in the model situated closer to this gender.
They also use analogy tasks to evaluate if a bias is present in a word embedding model. They check analogies by comparing pairs of word vectors. For example they search for the word complementing the puzzle: man is to doctor as woman is to \ldots? First they subtract the word ``man'' from the word ``woman''\footnote{This is pointwise subtraction of vectors, but for the simplicity we will often talk about words while really meaning vectors. What is meant should be clear from the context.} and then
they search for the ranked list of other words pairs that have similar vector difference. They also include in the formula a threshold to ensure that the resulting pairs could not be randomly picked.

However, as \protect\hyperlink{ref-Nissim2019Fair}{Nissim, Noord, \& Goot} (\protect\hyperlink{ref-Nissim2019Fair}{2019}) points out, there are some limitations to this approach. In practice, most of analogies implementations do not return any input words. This means that it does not make sense to expect the algorithm to return the same profession for both woman and man. Therefore, this method of bias detection seems limited. Other problems are related for example to the choice of pairs and words that are used to detect the presence of discrimination, as it is often subjective and without proper justification. Additionally, the choice of the parameter used in \protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) formula to ensure that word pairs are not picked by random, is also not justified and changing it drastically
influences the results.

\protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) touch upon the topic of biases regarding race and gender. They apply knowledge from well-known psychological studies such as Implicit Association Test to investigate the relation between human stereotypical thinking and model-learnt biases to discover a close relationship between these two. For the evaluation they use Word Embedding Association Test (WEAT) and the Word Embedding Factual Association Test (WEFAT).

\protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) propose a novel way of using cosine similarity to verify if the assumed resemblance between protected words and harmful attributes exists. They develop an approach that enables them to measure the bias for a class (such as gender, religion or race) and express the final result with a single metric. Their approach will be analyzed in details in the next chapter.

\protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) refer to Implicit Association Test (Greenwald et al., 1998) that measures the strength of associations between concepts or stereotypes by measuring human reaction time for special tasks. Humans naturally exhibit some biases which do not always cause social concern. One can imagine the intuitive associations between for example insects and flowers, and the feelings of pleasantness or unpleasantness. In general, people would rather associate flowers than insects with feeling pleasant, and this preference in a sense is a bias or
prejudice in some direction. However, this type of preference does not cause an uproar and is a rather morally neutral case. This example is rather used to show that the methodology used in IAT makes sense. \protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) were able to obtain similar results to the ones that measured reaction time with the use of WEAT metric. Unfortunately, further studies discover other biases and prejudices that directly influence the quality of other people's lives and therefore they should be taken care of.

@@

\hypertarget{cosine-similarity-and-distance}{%
\section{Cosine similarity and distance}\label{cosine-similarity-and-distance}}

@@

\begin{align} \tag{Sim}
\mathsf{cosineSimilarity}(A,B) & = \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert}
\end{align}
Cosine similarity is considered a proper tool for this operation as its result has a clear connection to geometry and at least for a low number of dimensions may be easily interpreted. Using this scale, one can compare vector similarities in a fairly clear manner. When the vectors are aligned perpendicularly to each other, their similarity equals 0 (which is the same as the cosine of 90 degrees). This tells us that the similarity between the vectors is small. As the angle between vectors decreases, cosine similarity approaches one, which stands for the greatest similarity. It is also possible to obtain negative cosine similarity. If the value approaches -1, this intuitively means that the words are contrary to each other.

@@

\hypertarget{cosine-distance-in-a-one-class-bias-detection}{%
\section{Cosine distance in a one-class bias detection}\label{cosine-distance-in-a-one-class-bias-detection}}

\protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) define similarity between words as the outcome inner product of their normalized vectors.
They focus on examining what the geometry of a word embedding is in regard to ``he'' and ``she'' words. In other words, whether the similarity between those concepts and other words reflects expected gender stereotypes. They test this hypothesis by investigating whether there is a connection between word embeddings representing certain professions and words referring to gender. They also evaluate whether automatically produced analogies between words reflect the stereotypes as well.

Here the gender bias of a word \(w\) is understood as its projection on the gender direction \(\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})\) (the gender direction is the top principal compontent of ten gender pair difference vectors). The underlying idea is that no bias is present if non-explicitly gender words are in equal distance to both elements in all explicitly gender pairs. Given the (ideally) gender neutral words \(N\) and the gender direction \(g\) the direct direct gender bias is defined as the average distance of the words in \(N\) from \(g\) (\(c\) is a parameter determining how strict we want to be):
\begin{align}
\mathsf{directBias_c(N,g)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},g)\vert^c}{\vert N \vert }
\end{align}

@@

\hypertarget{cosine-distance-in-a-multi-class-bias-detection}{%
\section{Cosine distance in a multi-class bias detection}\label{cosine-distance-in-a-multi-class-bias-detection}}

\protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) present a different approach towards finding similarities between classes of words. Cosine distance is used in the article as a measure to first argue for the existence of multi-class bias and then to show how through bias mitigation techniques the bias may be decreased.

They modify WEAT to a multi-class setting, introducing Mean Average Cosine similarity as a measure of bias (in fact, in the paper they report distances rather than similarities). Let \(T = \{t_1, \dots, t_k\}\) be a class of protected word embeddings, and let each \(A_j\in A\) be a set of attributes stereotypically associated with a protected word). Then:
\begin{align}
S(t_i, A_j) & = \frac{1}{\vert A_j\vert}\sum_{a\in A_j}\mathsf{cos}(t,a) \\
MAC(T,A) & = \frac{1}{\vert T \vert \,\vert A\vert}\sum_{t_i \in T }\sum_{A_j \in A} S(t_i,A_j)
\end{align}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{religion }\OtherTok{\textless{}{-}} \FunctionTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{]}
\FunctionTok{colnames}\NormalTok{(religion) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
                        \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
\NormalTok{religion}\SpecialCharTok{$}\NormalTok{wordClass }\OtherTok{\textless{}{-}} \FunctionTok{as.factor}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{wordClass)}
\FunctionTok{levels}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{wordClass) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
\FunctionTok{head}\NormalTok{(religion)  }\SpecialCharTok{\%\textgreater{}\%}  \FunctionTok{kable}\NormalTok{(}\AttributeTok{format =} \StringTok{"latex"}\NormalTok{,}\AttributeTok{booktabs=}\NormalTok{T,}
                      \AttributeTok{linesep =} \StringTok{""}\NormalTok{,  }\AttributeTok{escape =} \ConstantTok{FALSE}\NormalTok{, }
                      \AttributeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
                      \FunctionTok{kable\_styling}\NormalTok{(}\AttributeTok{latex\_options=}\FunctionTok{c}\NormalTok{(}\StringTok{"scale\_down"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

Assuming for a moment that the value of multi-class cosine distance is correct, one may question the interpretation. \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) summarize the averages of cosine distance per group (gender, race, religion). For now let us focus now on analyzing the values relating to religious biases. Here is the relevant fragment of table:

\begin{longtable}[]{@{}ll@{}}
\toprule
Religion Debiasing & MAC \\
\midrule
\endhead
Biased & 0.859 \\
Hard Debiased & 0.934 \\
Soft Debiased (\(\lambda\) = 0.2) & 0.894 \\
\bottomrule
\end{longtable}

@@

Using cosine similarity in high dimensions in word embeddings may also be prone to the curse of dimensionality. According to \protect\hyperlink{ref-Venkat2018Curse}{Venkat} (\protect\hyperlink{ref-Venkat2018Curse}{2018}) there are reasons to consider this phenomenon when searching for word similarities in higher dimensions. An experiment is conducted that aims at showing how the similarity values and variation change as the number of dimensions increases. The hypothesis made in the paper states that two things will happen as the number of dimensions increase. First, the effort required to measure cosine similarity will be greater, and two, the similarity between data will blur out and have less variation. The authors generate random points with increasing number of dimensions where each dimension of a data point is given a value between 0 and 1. Then they pick one vector at random from each dimension class and calculate the cosine similarity between the chosen vector and the rest of the data. Then they check how the variation of values changes as the number of dimensions increases. It seems like the more dimensions there are, the smaller the variance and therefore it is less obvious how to interpret the resulting cosine similarities. Maybe the scale should be adjusted to the number of dimensions and variance so that it still gives us sensible information about data. Yet, according to some articles cosine similarity in high dimensions is not reliable enough as it may be the case that choosing words at random may result in similar values as when picking them consciously.

@@

We will use the choice of protected words and stereotypical predicates used in \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}). This is a decent point of departure, as not only we want to compare our method to that of \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}), but also because this data format is fairly general (as contrasted, say, with a set up for binary stereotypes). Note also that the method we develop here can fairly easily be run for different stereotypization patterns. Let's start with explaining the method and its deployment using a dataset obtained for the religion-related protected words.

Let's load, clean a bit and inspect the head of the religion dataset we prepared. In order to obtain this dataset, we calculated the cosine distance between each protected word and each word from both the bias-related attribute groups, which were used in the original study, and to neutral and human control attributes which we added as control groups. For instance, for religion, the bias-related predicates (coming from the original study in \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019})) include muslim bias attributes, jew bias attributes, christian bias attributes (see a list in the Appendix).

@@

\caption{\label{tab:religionTableHead}Head of the religion dataset.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lllrrl}
\toprule
protectedWord & wordToCompare & wordClass & cosineDistance & cosineSimilarity & connection\\
\midrule
judaism & violent & muslim & 0.7141939 & 0.2858061 & different\\
judaism & terrorist & muslim & 0.7461333 & 0.2538667 & different\\
judaism & dirty & muslim & 1.2002599 & -0.2002599 & different\\
judaism & uneducated & muslim & 0.7885469 & 0.2114531 & different\\
judaism & greedy & jewish & 1.0026172 & -0.0026172 & associated\\
judaism & cheap & jewish & 1.2323229 & -0.2323229 & associated\\
\bottomrule
\end{tabular}}
\end{table}
\normalsize

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{religion}\SpecialCharTok{$}\NormalTok{con }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{connection)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: NAs introduced by coercion
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{religion}\SpecialCharTok{$}\NormalTok{pw }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{protectedWord)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: NAs introduced by coercion
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{religion}\SpecialCharTok{$}\NormalTok{pwFactor }\OtherTok{\textless{}{-}} \FunctionTok{factor}\NormalTok{(}\FunctionTok{paste0}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{protectedWord, religion}\SpecialCharTok{$}\NormalTok{connection))}
\NormalTok{religion}\SpecialCharTok{$}\NormalTok{pwIndex }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{pwFactor)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{source}\NormalTok{(}\StringTok{"../functions/cleanDataset.R"}\NormalTok{)}
\NormalTok{religion }\OtherTok{\textless{}{-}} \FunctionTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{]}
\NormalTok{religion }\OtherTok{\textless{}{-}} \FunctionTok{cleanDataset}\NormalTok{(religion,}\FunctionTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning in cleanDataset(religion, c("christian", "human", "jewish", "muslim", :
## NAs introduced by coercion
